---
title: "Lab 1: Exploratory Analysis of CEO Salary Data"
author: "Anamika Sinha, Asha Anju, Noah Randolph"
date: "June 3, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

An exploratory analysis of CEO and company data was conducted to explore CEO salary, with a particular interest in answering the question of whether company performance is related to CEO salary.

Exploratory data analysis techniques were used and will be presented with commentary on the findings. In general, CEO salary is treated as a response variable, while other datasets are treated as predictor variables.

```{r car}
library(car)
```

```{r load}
load("ceo_w203.RData")
```

## Begin mktval univariate analysis (REMOVE THIS HEADING)
Market value is commonly associated with company performance, which is of primary interest as it relates to CEO salary. Thus, we include an assessessment of its data in the variable, "mktval", univariately. First, a summary of the data: 

```{r summary}
summary(CEO$mktval)
```

Market value data shows to be in terms of float values, including negative numbers. The range is vast, with the maximum value far greater than the values in the interquartile range. In fact, the mean is greater than not just the median, but also the 3rd quartile. Since a stock price cannot go below $0, and market value for public companies is composed of the sum of the values of each individual stock, any negative values should be scrutinized.

```{r subset market value}
CEO$mktval[CEO$mktval < 0]
```

A subset of the market value data less than zero reveals that a coding error is apparent. The five values of "-1" likely represent data points that were unavailable for particular companies. Further analysis will be on a subset of the data excluding the five values of "-1".

```{r}
cleanmktval <- subset(CEO$mktval, CEO$mktval > 0)
```

Next, we create a histogram of the cleaned up market value data. The dataset is transformed by a factor of 1/10,000 to make for easier reading of the X-axis. The bins are set by the Freedman and Diaconis rule. We supply tick marks on the x-axis manually.
 
```{r market value histogram}
mktvalbillions <- cleanmktval/10000
hist(mktvalbillions, breaks = "FD", main = "Histogram of Market Value",
     xlab = "Market Value (end of 1990, in $ billions)", xaxt = "n")
axis(1, c(0, 0.5, 1.0, 1.5, 2.0, 2.5, 3.0, 3.5, 4.0, 4.5))
```

The histogram reveals the following characteristics:

1. Market value is skewed toward the positive and frequencies drop steeply between $0 and $300 million, for a decaying profile of frequencies as market value increases.

2. Although market value does not take on negative values, integer effects are not of concern due to the float type of data and its high dispersion.

3. Most of the market values fall below $600 million.

4. There is a large gap between highly dispersed data between roughly $2.7 billion and $4.3 billion.

Although the market values above $4 billion fall far outside of the bulk of the data, there is no strong evidence to conclude that they should be removed from the analysis as outliers.

## Begin ceoten univariate analysis (REMOVE THIS HEADING)

We next examine years of CEO tenure, which may relate to CEO salary as any normal salary typically grows with tenure. We manually set the cut points for our histogram to make 13 bins, which is roughly the square root of the number of data points.

```{r}
hist(CEO$ceoten, breaks = seq(0, 39, by=3), main = "Histogram of Years as CEO", xlab = "Years as CEO with Current Company", xaxt = "n")
axis(1, seq(0, 39, by=3))
```
The following observations are apparent in the histogram of CEO tenures:

1. The lower bound is at zero years (indeed, negative years would not make sense in this context).

2. The data is skewed positively, with more than one data point in the rightmost bin of 36 to 39 years.

3. CEO tenure frequencies peak in the lowest number of years and then mostly drop in a decaying form as the number of years increases.

4. There is a gap in tenures from 30 to 33 years and a decrease from 15 to 18 years, but these are most likely due to dispersion of lower frequencies of higher year values.


