---
title: 'Lab 1: Exploratory Analysis of CEO Salary Data'
author: "Anamika Sinha, Asha Anju, Noah Randolph"
date: "June 3, 2017"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

An exploratory analysis of CEO and company data was conducted to explore CEO salary, with a particular interest in answering the question of whether company performance is related to CEO salary.

Exploratory data analysis techniques were used and will be presented with commentary on the findings. We first explore data in univariate fashion, followed by multivariate analyses selected for their contribution to the exploration of CEO salary. In general, CEO salary is treated as a response variable, while other datasets are treated as predictor variables.

# Setup

We will utilize the "car: Companion to Applied Regression" package in R and load it, as well as the data file of interest, here.
```{r load}
library(car)
load("ceo_w203.RData")
```

# Exploratory Analysis

To get an initial understanding of the dataset, we'll look at the structure of its data frame and a summary of its variables.
```{r}
str(CEO)
summary(CEO)
```

Now we have an idea of the size of our dataset (184 observations) and the scale of each variable. The data appears to be all integers. We can see that college and grad appear to be variables of just 0 and 1 outcomes, so it will be useful to convert those to factors for labeling. Both profits and market value (mktval) have negative values.

Given that we are most interested in company performance and CEO salary, market value and profits appear to be our primary variables of interest, along with salary of course. The rest of which are secondary variables that could contribute to a better understanding of the primary relationships.

Given the large number of variables, scatterplot matrices are printed below to provide an overview of relationships between variables and identify areas of interest.

```{r}
scatterplotMatrix(~ CEO$salary + CEO$profits + CEO$mktval + CEO$age +
                          CEO$ceoten, data = CEO, diagonal = "histogram",
                  main = "Scatterplot Matrix 'A' for Key CEO Salary Variables")
```

In the upper diagonal half of the above matrix, it appears that the two company performance variables, profits and market value, both show a positive trend against CEO salary, although the data for both appear to be highly clustered at the low end. The secondary CEO age variable does not appear to have a strong influence on profits and market value, but there do appear to be slighter more high values to the right of the median age. The secondary variable CEO tenure (ceoten) looks to have higher values of profits and market value clustered at the low end of tenure. 

```{r}
scatterplotMatrix(~ CEO$salary + CEO$profits + CEO$mktval + CEO$comten,
                  data = CEO, diagonal = "histogram",
                  main = "Scatterplot Matrix 'B' for Key CEO Salary Variables")
```

In the upper diagonal half of the matrix above, the secondary variable company tenure (comten) appears to have slight positive relationships between profits and market value. Salary appears to be fairly evenly distributed across the span of company tenures.

## Univariate Analyses

We investigate the variables independently to understand their frequency distributions and to observe any anomalies, apparent data errors, and missing values. The dataset describing college attendance will not be analyzed due to the small number of CEOs who did not attend college (<4%), as can be seen below. 
```{r}
CEO$college <- factor(CEO$college, labels = c("Did Not Attend College",
                                              "Attended College"))
summary(CEO$college)
```


### Univariate Analysis of Salary Data

We start with the analysis of CEO Salary Data, our response variable of interest. We first examine the histogram of salary information.The individual salary amounts have been scaled to reflect the value in millions. We have set the break points manually but the bin width is same as that obtained using Freedman and Diaconis rule.

```{r}
hist(CEO$salary/1000, breaks = seq(0,6,.2),
     xaxt = "n",
     main = "Histogram of CEO Salary",
     xlab = "CEO Salary (million dollars)")
axis(1, seq(0,6,1))
```

Next we look at the summary statistics for the salary data.
```{r}
summary(CEO$salary/1000)
```

The following observations are apparent from our analysis of CEO salaries:

1. The distribution of CEO salaries is positively skewed with a peak around \$0.6 million.
2. Except the maximum value, all salaries are below \$3.0 million. Most of the salaries are below $2.0 million. 
3. The distance of maximum value from the 3rd quartile value is more than thrice the interquartile range, thus making the data point an outlier. 

Eventhough the maximum salary is an outlier, there is no strong evidence to conclude that it should be removed from our analysis.

### Univariate Analysis of Market Value

Market value is commonly associated with company performance, which is of primary interest as it relates to CEO salary. Thus, we include an assessessment of its data in the variable, "mktval", univariately. First, a summary of the data: 

```{r summary}
summary(CEO$mktval)
```

The market value range is vast, with the maximum value far greater than the values in the interquartile range. In fact, the mean is greater than not just the median, but also the 3rd quartile. Since a stock price cannot go below $0, and market value for public companies is composed of the sum of the values of each individual stock, any negative values should be scrutinized.

```{r subset market value}
CEO[CEO$mktval < 0, ]
```

A subset of the data less than zero reveals that a coding error is apparent. The -1 values in both market value and profits likely represent data points that were unavailable for particular companies. Further analyses in this report will be on subsets of the data excluding the values of -1 for market value and profits.

```{r}
cleanCEO <- subset(CEO, CEO$mktval > 0)
```

Next, we create a histogram of the cleaned up market value data. The dataset is transformed by a factor of 1/1,000 to make for easier reading of the X-axis. The bins are set with even widths at a number of 20, which is an amount between two common rules, one of which is the square root of the number of data points, and the other is the Freedman and Diaconis rule.
 
```{r market value histogram}
mktvalbillions <- cleanCEO$mktval/1000
hist(mktvalbillions, breaks = seq(0, 50, 2.5), 
     main = "Histogram of Market Value",
     xlab = "Market Value, end of 1990 (billion dollars)")
```

The histogram reveals the following characteristics:

1. Market value is skewed toward the positive and frequencies drop steeply between \$250 million and $500 million, for a decaying profile of frequencies as market value increases.

2. Although market value does not take on negative values, integer effects are not of concern due to its high dispersion.

3. Most of the market values fall below $500 million.

4. There is a large gap in data between roughly \$27.5 billion and $42.5 billion.

Although the market values between \$4 billion and $5 billion fall far outside of the bulk of the data, there is no strong evidence to conclude that they should be removed from the analysis as outliers.

### Univariate Analysis of Company profits

Next, we investigate company profits as it is also an indicator of company performance.

```{r}
summary(cleanCEO$profits)
```

We see that there are negative profits but the number of companies with 
negative profits may be low as the first quartile value is still positive. There is a wide range in values with the maximum profit going as high as 13 times the 3rd quartile value. The mean being much higher than the median and even more than the third quartile values indicates the presence of a few very high profit values.

For the purpose of further analysis, we will use the cleanCEO dataset created and explained above in market value analysis. (5 values with coding errors have been dropped as the value of profit is also -1 along with mkatval being -1) 


```{r}
hist(cleanCEO$profits,
     breaks = 20,
     ylim = c(0,175),
     main = "Profits for Companies in 1990",
     xlab = "Profits(million dollars)")
```



 
 Observations from company profit histogram: 

1. Most companies have profits below $250 million. 

2. The histogram has a slight positive skew due to the presence of some outliers. They seem to be pushing the mean to even beyond the 3rd quartile. The mean is much higher than the median value.

3. Since most of the profits are from \$0 to $250 million, that bin is worth some further analysis. We split this bucket later to see if there are any apparent trends.

4. It will be interesting to see the salaries of CEOs of the 10 companies with negative profits.

### Univariate Analysis of CEO Age

We next examine the age of CEOs to understand the distribution and its features.

```{r}
summary(CEO$age)
```

```{r}
hist(CEO$age, breaks = "FD",
     main = "Histogram of CEO Age",
     xlab = "Age of CEO (years)")
```

The following observations are apparent from our analysis of age information:

1. The ages cover a large range with a minimum age of 21 years and maximum age of 86 years.
2. The distribution peaks peak around 55-60 years of age and have a slight negative skew. 
3. 50% of the CEOs in the dataset have ages in a 10 year range between 51 and 61.25 years.

### Univariate Analysis of CEO Tenure

We next examine years of CEO tenure, which may relate to CEO salary as any normal salary typically grows with tenure. We manually set the cut points for our histogram to make 13 bins, which is roughly the square root of the number of data points.

```{r}
hist(CEO$ceoten, breaks = seq(0, 39, by=3), 
     main = "Histogram of Years as CEO", 
     xlab = "Years as CEO with Current Company", 
     xaxt = "n")
axis(1, seq(0, 39, by=3))
```

The following observations are apparent in the histogram of CEO tenures:

1. The lower bound is at zero years (indeed, negative years would not make sense in this context).

2. The data is skewed positively, with more than one data point in the rightmost bin of 36 to 39 years.

3. CEO tenure frequencies peak in the lowest number of years and then mostly drop in a decaying form as the number of years increases.

4. There is a gap in tenures from 30 to 33 years and a decrease from 15 to 18 years, but these are most likely due to dispersion of lower frequencies of higher year values.



### Univariate analysis of Company tenure


We analyzed the total number of years CEOs stayed with the 
company including years before or after becoming CEO.

```{r}
summary(CEO$comten)
```

There is a wide range in the number of years CEOs have stayed with the company.
The mean and median are almost equal which indicates a uniform distribution of 
data.

```{r}
hist(CEO$comten, 
     ylim = c(0,40),
     main = "Histogram of Years in Company",
     xlab = "Number of Years in Company"
     )
```

The following observations are apparent: 

1. We see that there is almost a uniform spread of data across the number of years from zero to forty years. 

2. There is a small spike at 30-35 years and 1 observation in the 55 to 60 years. 


### Univariate analysis of CEOs who attended graduate school

Since graduate school (grad) is a categorical variable, we create a factor variable
and assign levels of "Attended" and "Did Not Attend" to it.

```{r}
CEO$grad <- factor(CEO$grad,
                      labels = c("Did Not Attend", "Attended"))
```

```{r}
plot(CEO$grad, ylab = "Count of CEOs",
     main = "CEO Graduate School Attendance")
```

```{r}
summary(CEO$grad)
```


Observations:

1. We see that the number of CEOs without a graduate degree is pretty comparable to the number of CEOs with graduate degree.

2. This means that the data is distributed uniformly between the 2 levels.

## Multivariate Analyses

Now that we understand the type and general nature of the data in each variable, we turn to multivariate analyses to begin to answer the question of whether CEO salary is related to company performace. We are using market value and profits as proxies for company performance for the primary assessment of salary as it relates to company performance.

### Multivariate Analysis of Market Value Versus CEO Salary

Market value is plotted against CEO salary as it is one measure of company performance. We are interested in seeing the degree to which the level of CEO salary is related to the market value of that CEO's company.

Because of the high clustering of market value data seen in the scatterplot matrix of the initial exploration above, market value will be plotted using a $log_{10}$ scale. We exclude observations (whole rows of data) including the -1 values found in the univariate analysis of market values (though they would not plot on a log scale even if included).

```{r}
plot(log10(cleanCEO$mktval), cleanCEO$salary/1000, main = "Market Value Versus CEO Salary", 
     xlab = expression(paste(Log[10], 
                             " of Market Value, end of 1990 (million dollars)")), 
     ylab = "Salary (million dollars)")
abline(lm(cleanCEO$salary/1000 ~ log10(cleanCEO$mktval)))
```

The plot above, with its linear regression line, shows a clear positive relationship between salary and the $log_{10}$ of market value. The correlation, below, also shows that indeed some relation exists. However, the relationship cannot be said to be linear, since market value was plotted in $log_{10}$ scale. 
```{r}
cor(cleanCEO$mktval, cleanCEO$salary)
```

# REVISIT THIS SECTION****

We group the data into different bins using market values and analyse the salary statistics for these bins. This will help

```{r}
#mktval_bins <- cut(cleanCEO$mktval/1000,
#                        c(0,1,2,5,10,20,60))
#summary(mktval_bins)
```

```{r}
#salary_median_by_mktval = by(cleanCEO$salary/1000, mktval_bins, median)

#boxplot(cleanCEO$salary ~ mktval_bins,
#     main = "Plot of median salary by market value",
#     ylab = "Median Salary in $ millions",
#     xlab = "Marketvalue in $ billions")
```
These plots further strengthen our assessment of the positive relationship between salary and market values (proxy for company performance).

### Bivariate Analysis of Salary and Profit

Profit is another indicator of company performance. We shall explore if there 
is a relationship between company profit and CEO salaries.

Seeing the high clustering of profit data in the scatterplot matrix, we were tempted to plot the log of profits against salary to look for a relationship. However, due to the presence of 10 negative profit values, we decided against log values as they remove negative profits from the analysis.

Instead, we grouped the profit data to see if salaries for CEOs showed a trend across profit bins.

Histogram analysis of profits done earlier led us to create a bin for negative profits. We created two more bins of profits in the ranges of \$0 to \$125 million and \$125 to $250 million. Finally, we created another bin of values greater than \$250 million to bundle higher values of profit together.

```{r}
profit_bin = cut(cleanCEO$profits, breaks = c(-500,0,125,250, Inf), labels = 
      c("Profit < 0", "0 < Profit < 125","125 < Profit < 250", "250 < Profit"))
summary(profit_bin)
```

```{r}
boxplot(salary/1000 ~ profit_bin, data = cleanCEO, cex.axis = .7,
        main = "CEO Salaries by Company Profit",
        ylab = "Salary (million dollars)",
        xlab = "Profit (million dollars)")
```
From the graph above, we notice an upward trend in median salary with increasing profits in the realm of positive profits . There are some outliers with high salaries in the zero to $250 million bucket. The wider range of salaries in the \$0 to \$250 million as well as greater than \$250 million bucket is also noticeable. 

It is also interesting to note that CEOs in companies with negative profits did not get paid the lowest. Infact, the median value in that bucket is slightly higher than the \$0 to $250 million bucket.

```{r}
cor(cleanCEO$salary, cleanCEO$profits, use = "complete.obs")
```
The correlation value of 0.39 above does show a somewhat positive correlation.

## Analysis of Secondary Variables

In this section, we analyse the impact of the remaining variables on our primary variables of interest viz. salary, market value and profits.

### Effects on CEO Salary

We look at the influence of the secondary variables on the CEO salary.

#### CEO Tenure

The variable ceoten captures the number of years of service as CEO. The data is reported as integers and ranges from  0 to 37. 75% of the values are concentrated in the first one third of the overall tenure spread. This clustering makes scatter plots less informative. Therefore to analyse the influence of ceo tenure on the primary variables, we group them into bins. Fifty percent of the tenure values are below 6 years and therefore we create two bins in this range to better understand the data. 

```{r}
ceoten_bins <- cut(cleanCEO$ceoten,
                        c(0,3,6,10,20,40),
                        include.lowest = TRUE,right = FALSE, 
                        labels = c("<3 yrs","3-6 yrs", "6-10 yrs", 
                                   '10-20 yrs','20-40 yrs'))

# Gives no of datapoints in each bin
summary(ceoten_bins)
```

We create a boxplot of salary for the various tenure groups. The y axis has been limited to salary of \$2.5 millions.  
```{r}
boxplot(cleanCEO$salary/1000~ ceoten_bins,
     main = "Plot of CEO Salary by CEO Tenure",
     xlab = "CEO Tenure (years)",
     ylab = "Salary (million dollars)",
     ylim = c(0,2.5))

```
The median salaries for CEOs appear to increase with tenure for tenure less than 10 years. Beyond that the effect of years of tenure seem to reduce.

#### Company Tenure

The variable comten captures the number of years of service with the company. The data is fairly uniformly distributed across the years. To simplify our analysis, we group the data based on years of tenure at the company. We have created bins of size 10 years except for the last bin which captures data for a twenty year period. This is because of the low frequency of data with such high tenure. 

```{r}
comten_bins <- cut(cleanCEO$comten,
                   c(0,10,20,30,40,60),
                   labels = c('<10', '10-20 yrs','20-30 yrs',
                                   '30-40 yrs', '40-60 yrs'),
                   right = FALSE)

# Gives no of datapoints in each bin
summary(comten_bins)
```

We create a boxplot of salary for the various tenure groups. The y axis has been limited to salary of \$2.5 millions. 

```{r}
boxplot(cleanCEO$salary/1000~ comten_bins,
     main = "Plot of CEO Salary by Company Tenure",
     xlab = "Company Tenure (years)",
     ylab = "Salary (million dollars)",
     ylim = c(0,2.5))

```

The median salary and its distribution appears fairly similar for tenure less than 40 years. The low median salary for the highest tenure group comes as a surprise.

#### Age

Age data is clusterd with nearly 50% of the data spread on a 10 year period. This makes drawing conclusions using scatterplot difficult. Therefore we apply grouping to age data as well.  

```{r}
age_bins <- cut(cleanCEO$age, 
                     c(0,40,50,60,70,90),
                labels = c('<40 yrs', '40-50 yrs', '50-60 yrs',
                           '60-70 yrs', '70-90 yrs'),
                     right = FALSE)

# Gives number of datapoints in each age range
summary(age_bins)
```

We create a boxplot of salary for the various tenure groups. The y axis has been limited to salary of \$2.5 millions. 

```{r}
boxplot(cleanCEO$salary/1000~ age_bins,
     main = "Plot of CEO Salary by Age",
     xlab = "Age of CEO (years)",
     ylab = "Salary (million dollars)",
     ylim = c(0,2.5))

```

For age in range 40 to 70 years, the median salary shows an increasing trend with age. The lower median salary and greater spread of salary for the highest age bracket is surprising.



### Effects on Market Value

We analyse the impact of secondary variables on market value. Due to the large range of market values and clustering at the lower level values, we perform grouping on the market value variable. Due to the shape of the market value distribution we have more bins at the lower end of the market value spectrum. 

```{r}
mktval_bins <- cut(cleanCEO$mktval/1000,
                        c(0,1,2,5,10,20,60),
                   labels = c('<1','1-2','2-5','5-10','10-20','20-60'))
summary(mktval_bins)

```

#### CEO Tenure

There does not seem to be a significant difference in the median salaries of CEOs who attended college compared to those who did not.

```{r}
boxplot(cleanCEO$ceoten ~ mktval_bins,
     main = "Plot of CEO Tenure by Market Value",
     ylab = "CEO Tenure (years)",
     xlab = "Marketvalue (billion dollars)")

```

The median CEO tenure tend to be decreasing with increase in market value of company. But there is an increase in the tenure for companies in \$10 billion to $20 billion range.

#### Company Tenure

```{r}
boxplot(cleanCEO$comten ~ mktval_bins,
     main = "Plot of Company Tenure by Market Value",
     ylab = "Company Tenure (years)",
     xlab = "Marketvalue (billion dollars")
```
The companies in different market value ranges tend to behave slightly differently. For companies with a market value below \$2 billion, the median tenure is around 20 years. For companies with market values in \$2 billion to $10 billion, the median tenure is around 26 years. The median company tenure is closer to 30 years for companie with market values above \$10 billion.

#### Company Age

```{r}
boxplot(cleanCEO$age ~ mktval_bins,
     main = "Plot of CEO Age by Market Value",
     ylab = "Age of CEO (years)",
     xlab = "Marketvalue (billion dollars")
```
The median age of CEOs appear to increase with increase in market value of the company. The exception is out group with the largest market value, which has the lowest median age.


## Confounding Effects

We have used profits and market values as proxy for company performance. We have explored the relationship between these individual variables and salary. However the two variables themselves exhibit a positive relationship.

```{r}
boxplot(cleanCEO$profits ~ mktval_bins,
        main = "Plot of Company Profits vs Market Value",
        xlab = "Market Value (billion dollars)",
        ylab = "Company Profit (million dollars)")
```
Therefore a part of the positive relationship of market value with Salary may be explained by profits.

```{r}
boxplot(age ~ profit_bin, data = cleanCEO, cex.axis = .7,
        main = "CEO Age by company Profit buckets",
        ylab = "Age in years",
        xlab = "Profits(millions dollars)"
        )
```

The graph above shows a very slight increasing trend in median age of CEOs with profit when profit is positive. But since most of our profits are in the 0 to 125 million range, the wide variablity of in this range age cannot be ignored. Overall, age does not seem to have a confounding effect on company profits
The slight correlation is confirmed by the value below.
```{r}
cor(cleanCEO$age, cleanCEO$profits, use = "complete.obs")

```

```{r}

boxplot(comten ~ profit_bin, data = cleanCEO, cex.axis = .7,
        main = "Company tenure of CEO by company Profit buckets",
        ylab = "company tenure in years",
        xlab = "Profits(millions dollars)")
```

From the graph above, we do see an increasing trend in median CEOcompany tenure as company profit increases. This is  prominent in companies with very high profits. One could argue that the number of years the CEO has been in a company does have a relationship with very high company profit.Also interesting is that companies with negative profits have lowest median company tenure.

```{r}
cor(cleanCEO$comten, cleanCEO$profits, use = "complete.obs")

```

```{r}

boxplot(ceoten ~ profit_bin, data = cleanCEO, cex.axis = .7,
        main = "CEO tenure by company Profit buckets",
        ylab = "ceo tenure in years",
        xlab = "Profits(millions dollars)")
```
From the graph above, median ceo tenure seems to have a very slight decreasing trend with profit increase. It is interesting to note that companies with high profits have the lowest median CEO tenure, even a bit lower than compnies with negative profits. This may indicate a slightly high CEO turnover at companies with higher profits.
```{r}
cor(cleanCEO$ceoten, cleanCEO$profits, use = "complete.obs")

```
### Missing values, coding errors etc.

For two data points, the company tenure is less than the CEO tenure. Since these two data points did not change the results of our analysis, we have left them in our dataset. 
